### 2020-07-06 response/response-demography-3.tex

\section*{Deputy Editor}

\begin{enumerate}

\item The overall topic and research question is not clear in the
introduction. The main topic relates to changes in birth spacing over
time, which are influenced by sex selection/sex composition and maternal
education, and how this affects fertility and infant mortality. This is
a long list of interesting questions and analyses but this is confusing
for the reader. It would be useful to have a succinct statement about
what the study is about (what are key independent variables/concepts and
key outcomes) and place that in the introduction and before the data and
analyses section. This way, the reader can understand why the analyses
will proceed in a particular order and what is being tested. In
addition, the titles of the figures and tables should be reviewed to see
if they could better reflect the analyses undertaken. Overall, the
reader needs to be reminded at the beginning and in each section what
you aim to do.

I have rewritten the Introduction to emphasize the main question I address
(``how did birth spacing respond to these changes, especially the introduction and spread 
of sex selection?'') and the two questions addressing the effects of changes in 
birth spacing (``to what extent do changes in birth spacing bias our standard fertility 
estimates?'' and `` how do the changes in birth spacing and the increased use of abortions 
impact children's chances of survival?'').

@@

I have rewritten the Introduction to emphasize the main question I address
(``how did birth spacing respond to these changes, especially the introduction and spread 
of sex selection?'') and the two questions addressing the effects of changes in 
birth spacing (``to what extent do changes in birth spacing bias our standard fertility 
estimates?'' and `` how do the changes in birth spacing and the increased use of abortions 
impact children's chances of survival?'').

Following the three questions, I now outline how the study progresses focusing on 
how I examine each of the three questions, including the key outcomes and key explanatory 
variables.
To prevent exceeding the word count I have also edited the entire introduction.
Although sligthly longer than original, it is still less than 900 words 
(not counting references).

I have edited the introduction of each section to highlight what I question
I am addressing.
Furthermore, I have removed some of the repetition in, for example, the
discussion of tempo effects and mortality, so that the motivation is now only
in the Introduction rather than being spread over the Introduction and the
relevant sections.

Finally, I have made the discussion of the distribution of spacing a subsection
of the spacing section and combined the discussion of the results to another new 
subsection as well.
I hope this will make it easier for the reader to see the answer to the main
question of the paper.

@@

[Response:] I now define ``sex selection'' as ``the selective abortion of female fetuses 
based on prenatal sex determination'' in the second paragraph of the paper.

\item Please provide a motivation for using maternal education as a key
stratification measure and justify education cutoffs (which have meaning
in India).

[Response:]

To address this, and the questions about theories of female labor supply below, I have
combined and reorganized the background and conceptual framework sections.
I begin this combined section with a discussion of the three main reasons for using 
maternal education as a key stratification measure.
First, higher female education is associated with lower fertility, and, thereby, with
potential increased use of sex selection.
Second, female labor force participation in India, as well as other developing countries, 
tend to first decrease and then increase with education.
Finally, child morbidity and mortality decreases the better educated the mother.

After the discussion of why female education is a key variable, I explain the choice
of cut-offs, which is based on the prior literature on fertility and son preference in 
India and the NFHS reports.
The exception to the the cut-offs in the NFHS reports is that I combine
the <5 years and 5--7 years of schooling completed and 
the 8--9 and 10--11 years of schooling completed to reach sufficient sample sizes.

This grouping also allows me to capture the U-shape in female labor force participation.
The upper level of education could be split into having graduated secondary and having 
graduated with a university degree since women with a completed university degree have 
a higher labor force participation, but the sample of university graduates is too small 
to allow for estimations, especially for the first three rounds of the surveys.
Furthermore, with four groups there is already a substantially number of results to 
present.

@@

[Response:]

\item I see the point to reviewer’s comment #4 about the discussion on
women’s labor supply. It’s a complex topic to introduce in a few brief
sentences. Similarly, referring to Goldin’s U-shaped theory without a
more detailed explanation could be confusing for those not well versed
in this topic. Perhaps you could also mention that decreasing FLFP is
unexpected for developing countries and some potential reasons for this
in India.

[Response:]

I agree, and to help address this, and the question about maternal education cut-offs
above, I have combined and reorganized the conceptual framework and background sections.

The main relevant theory on female labor force participation and education is the income 
and substitution effects discussed in \citet{Hotz1997} and \citet{schultz97}.
With the reorganization of the sections, the introduction of this theory now more naturally 
follows the discussion of increasing education in India.
I then discuss the declining female labor force participation and that it may come
about because of the rapid income increases.

Since the Goldin theory is not central to this argument, I have removed the sentence 
on the theory.
It is worth pointing out, however, that a decline in female labor force participation  
is in line with the hypothesis of a U-shaped labor force participation as a country 
develops \citep{Goldin1994}.
What is ``surprising" for India is that we have not yet seen any evidence of the upward 
sloping part.
I do think that the combination of rapidly rising male income, son preference, and 
relatively meager job opportunities for women with, say, high school education have a lot 
to do with this.

@@

[Response:]

\item In the context of the paper, I don’t see the point of the
discussion relating to labour supply.

[Response:]

The discussion of labor supply was disjointed and unclear because I had split the 
background and conceptual framework sections, making it difficult to see the connection.
As explained above, I have combined and reorganized those two sections into one and
removed the Goldin theory discussion since it is not directly relevant to my main
argument of why labor supply is relevant.

### 2020-07-06 paper/sex_selection_spacing-ver4.tex

\section{Female Education and Labor Force Participation in India}

To set the stage for the subsequent analyses, I first describe how female education
and labor force participation has changed.
Both are potentially important factors in the birth spacing decision, as I discuss
in the next section.

Female educational attainment has increased substantially over time.
Figure \ref{fig:education_over_time} shows the distribution of schooling by birth 
cohort for urban and rural women, 
twenty years or older, whether married or not, based on the four rounds of the NFHS.
The education groups are no education, one to seven years, and eight to eleven years,
and twelve years and above.

@@

For rural areas, the percentage of women with no education has gone from around 90 percent
for the 1930s cohorts to less than 20 percent for the 1990s cohorts. 
The proportion of women with one to seven years of education has remained remarkably
constant at around ten percent.
The difference is made up by the women with eight or more years of education, who have
gone from almost zero for the 1930 cohort to more than sixty percent for the 1990s cohorts,
with about half in the eight to eleven years group and 
the other half in the twelve years or more group.

Female education is higher in urban areas than in rural areas.
Around sixty percent of urban women born in the 1930s had no education, 25 percent had
between one and seven years, about ten percent had eight or more years of education, and 
only five percent has twelve years or more.
The proportions with no education and one to seven years have both declined to just below 
ten percent for the latest cohort.
Although the proportion of women with eight to eleven years of education has increased
to about twenty, most of the increase in urban female education has come from the 
twelve plus group, which now account for more than half of all urban women.

Even as the level of female education has increased, the female labor force participation 
in both urban and rural areas has stagnated or decreased
\citep{Klasen2015,Fletcher2017,Afridi2018,Bhargava2018,Chatterjee2018,Bhargava2019}.
A decline in female labor force participation at the beginning of development is
consistent with the hypothesis of a U-shaped labor force participation as a country
develops \citep{Goldin1994}.
India's female labor force participation is, however, lower than most other
countries and more in line with countries in the Middle East and North Africa, and
does not yet show any signs of increasing
\citep{Klasen2015,Chatterjee2018}.

The NFHS data show a U-shaped relationship between education and working for married
women, with the highest percent working for women with either no education 
or with twelve or more years and the lowest for women with eight to eleven years of 
education.%
\footnote{
See Appendix Figures \ref{fig:work_by_survey} through \ref{fig:work_family_by_survey}. 
}
Women are more likely to report working if they live in rural than urban areas 
and the older they are.
For married women, the percentage currently working changes relatively little over time. 
All education and age groups have, however, become substantially more likely to work for 
a family member.

The increases in female educational attainment imply that access to education has 
expanded beyond the higher castes. 
One possible effect of the rapid expansion in female education, and the 
associated change in the composition of better-educated women, is that the behavior
of the better-educated would change. 
However, ``Sanskritization'' implies that as lower castes females gain access to 
education and their husbands' income increases, the women adopt higher caste 
norms such as stronger son preference and a retraction from the formal labor 
market \citep{Srinivas1956,Chen1995,Abraham2013,Chatterjee2018}.
The low and declining female labor force participation is an indication that this
process still operates.

\section{Birth Spacing: Mechanisms and Prior Findings}

To provide a framework for understanding how birth spacing responded to the advent 
of sex selection and the substantial changes in income, education, and labor force 
participation, I next discuss relevant theories and prior findings on birth spacing.

The standard economic argument for the predominant association between
increasing female education and shorter spacing is that parents incur
time costs when they have children \citep{Hotz1997,schultz97}.
Specifically, if having children require the mother to reduce her market work, parents 
can lower the cost of children by shortening birth spacing to 
take advantage of economies of scale in childrearing \citep{Vijverberg1982}.

The low and declining female labor force participation, especially for younger women, 
suggests, however, that families face little incentive to space children more closely 
together for economic reasons.
One explanation is that household income has increased so substantially that the income 
effect dominates any substitution effect.
Two findings speak for this effect.
First, although real wages for both men and women have almost doubled between 1987 and 
2011, the mean male wage is still close to 70 percent higher than the female wage 
\citep{Klasen2015,Bhargava2018}.
Second, women’s labor supply appears to be more negatively elastic to their husbands' 
wages than it is positively elastic to their wages \citep{Bhargava2018}.

The combination of rising incomes and continued son preference may lead to even longer 
spacing than the standard income effect alone predicts.
As women's education increases, their productivity in the production of offspring human 
capital also increases.
With relatively more boys born because of increased access to sex-selective 
abortions and an increasing potential income for (male) offspring, demand for 
better-educated women can increase, even if they do not participate in the labor market 
\citep{Behrman1999}.
If more and ``better'' parental attention per child results in higher child ``quality,'' 
we should expect longer birth intervals 
\citep{Zajonc1975,Zajonc1976,Razin1980}.
The evidence on spacing's effect on child quality measures such as IQ 
and education is, however, mixed for developed countries and non-existing 
for developing countries
\citep{Powell1993,Pettersson-Lidbom2009,Buckles2012,Barclay2017}.
The exception is one crucial aspect of child quality---health and mortality---where
longer spacing does lead to better outcomes, although this relationship weakens with 
maternal education \citep{Whitworth2002,Conde-Agudelo2012,Molitoris2019}.

% While one might expect that increasing income and female education would also lengthen
% average birth spacing through better access and knowledge of contraception, the empirical 
% relationship between contraception use and birth spacing is ambiguous 
% \citep{Tulasidhar1993,Whitworth2002,Bhalotra2008,Yeakey2009,Kim2010,Soest2018}.
% Furthermore, the ability to successfully use ``low efficacy'' contraceptive methods, 
% such as the rhythm method, is increasing in education, while there is no difference 
% across education levels when it comes to modern contraceptives such as pills or 
% injectables \citep{Rosenzweig1989}.

The introduction of sex selection allows parents to avoid the birth of girls
but increases the expected duration to the next birth.
Theory suggests that sex selection increases with lower desired fertility, and, for a 
given desired number of children, the higher the parity \citep{Portner2015b}.
The increased use of sex selection in India with education and in urban
areas is consistent with a lower desired fertility for both groups
\citep{das_gupta97,retherford03b,Guilmoto2009a,Portner2015b,Jayachandran2017}.
Better-educated and urban women also tend to live in households with higher income, which 
better enables them to access prenatal sex determination and lowers the relative costs of 
using sex selection and have long birth intervals.

We may even observe a ``delay'' effect, where parents conceive later because they know 
that short spacing is detrimental to the next child's health and that the next child is 
more likely to be a boy because of access to sex selection.
Working in the opposite direction is that women with more education can, in principle, 
space their children closer together without substantially increasing child mortality 
risk.
Given the apparent lack of pressure to return to the labor force,
this effect is, however, likely to be small.

In summary, with substantial increases in husbands' income and a declining female labor 
force participation, I expect a push toward longer birth spacing over time, independent
of education levels, based on the income effects and the effects of spacing
on child outcomes.
As desired fertility decreases with increasing education, I, furthermore, expect 
birth spacing to increase the most among the better-educated because their household
income increases the most---even with declining female labor participation---and 
their use of sex selection.
Even with the substantial increase in the number of better-educated women, 
"Sanskritization" implies that the changing composition does not substantially change 
this group's use of sex selection.

@@

Direct information on the use of sex selection is not available, so I compare different 
periods, based on the changes in access and legality of prenatal sex 
determination in India.
Abortion has been legal in India since 1971.
Reports of sex determination appeared around 1982--83, and the number of clinics 
quickly increased \citep{Sudha1999,bhat06,Grover2006}.
In 1994, the Prenatal Diagnostic Techniques (PNDT) Act made determining and communicating 
the sex of a fetus illegal.%
\footnote{
Details about the act are at \href{http://pndt.gov.in/}{http://pndt.gov.in/}.
There is little evidence that the ban significantly affected sex ratios \citep{Das-Gupta2016}.
}
Although the use of sex selection increased even after 1994, we may have passed a 
turning point in its use \citep{Das_Gupta2009,Diamond-Smith2015}.
I, therefore, use four periods: 1972--1984, 1985--1994, 1995--2004, and 2005--2016.
The allocation of spells into periods is determined by when conception, and, therefore, 
decisions on sex selection, can begin. 
Hence, some spells cover two periods, which may bias downward the differences between the 
periods.

@@

Women with different education levels have different hazard profiles
\citep{Whitworth2002,Bhalotra2008,Kim2010}.
Furthermore, as discussed above, the use of sex selection increases with education.
I, therefore, divide women into four educationgroups: no
education, one to seven years of education, eight to eleven years of education,
and twelve and more years of education.
The models are estimated separately for each group to reduce any
potential problem from including other variables as proportional.

I capture sex composition with dummy variables for the
possible combinations, ignoring the ordering of births.
Area of residence is a dummy variable for living in an urban area.

>>>>

% Combine descriptive and theory sections into one

\section{Background and Conceptual Framework}

To set the stage for the subsequent analyses and provide a conceptual framework for 
understanding birth spacing, I first discuss female education and labor force 
participation in India and relevant theories on birth spacing.

% [Female education]

Female education is a crucial explanatory variable here for three reasons.
First, higher female education is associated with lower fertility and increased use of 
sex selection
\citep{das_gupta97,dreze01,bhat03,retherford03b,Guilmoto2009a,Portner2015b,Jayachandran2017}.
Second, female labor force participation in India, as in other developing countries, 
first decreases and then increases with education
\citep{Klasen2015,Fletcher2017,Afridi2018,Bhargava2018,Chatterjee2018,Bhargava2019}.
Finally, child mortality decreases with maternal education
\citep{Rosenzweig1982a,Whitworth2002,Maitra2008}.

[TK where to put son preference discussion?]

I divide eduction into four groups: no education, one to seven years, eight to eleven 
years, and twelve and more years.
The latter two correspond to having completed primary and completed secondary school, 
respectively.%
\footnote{
Although there are variations by state, elementary education in India consists of a 
primary school covering grades one through five and an upper primary---or middle 
school---covering grades six through eight. 
Similarly, secondary education covers grades nine and tenth for ``secondary education'' 
and eleven and twelve for ``upper secondary.''
}
To ensure that the results are comparable with the prior literature on fertility
and mortality in India, I follow the NFHS reports, except that I combine the 
less than five years and five to sevens years of schooling completed and the eight to nine 
and ten to eleven years of schooling completed.
There is a trade-off between having groups large enough for the analyses below and having 
groups detailed enough to capture differences in behavior across education levels, such as 
the U-shaped relationship between education and labor force participation.

% [what has happened to education]
Female education has increased substantially over time in rural and urban areas but 
is still substantially higher in urban than in rural areas.
Figure \ref{fig:education_over_time} shows the distribution of schooling by birth 
cohort for urban and rural women, 
twenty years or older, whether married or not, based on the four rounds of the NFHS.
In rural areas, women with no education have gone from 90 percent for the 1930s cohorts to 
less than 20 percent for the 1990s cohorts.
Women with eight or more years of education have gone from almost zero for the 1930 cohort 
to more than sixty percent for the 1990s cohorts, with about half in the eight to eleven 
group and the other half in the twelve plus group.
In urban areas, the proportions with no education or one to seven years have declined 
to just below ten percent each.
Most of the increase in urban female education came from the twelve plus group, which 
now account for more than half of all urban women for the lastest cohorts.

@@

% [female labor force participation]

The standard economic argument for spacing becoming shorter with increasing female education 
is that parents incur time costs when they have children \citep{Hotz1997,schultz97}.
Specifically, if having children require the mother to reduce her market work, parents 
can lower the cost of children by shortening birth spacing to 
take advantage of economies of scale in childrearing \citep{Vijverberg1982}.

However, even as the level of female education has increased, the female labor force 
participation in both urban and rural areas has stagnated or decreased
\citep{Klasen2015,Fletcher2017,Afridi2018,Bhargava2018,Chatterjee2018,Bhargava2019}.
India's female labor force participation is now lower than most other countries and does 
not yet show any signs of increasing \citep{Klasen2015,Chatterjee2018}.
In line with previous research, the NFHS data show a U-shaped relationship between 
education and working for married women, with the highest percent working for women with 
either no education or twelve or more years and the lowest for women with eight to 
eleven years of education.%
\footnote{
See Appendix Figures \ref{fig:work_by_survey} through \ref{fig:work_family_by_survey}. 
}

The low and declining female labor force participation, especially for younger women, 
suggests that families face little incentive to space children more closely 
together for economic reasons.
One explanation is that household income has increased so substantially that the income 
effect dominates any substitution effect.
Two findings speak for this effect.
First, although real wages for both men and women have almost doubled between 1987 and 
2011, the mean male wage is still close to 70 percent higher than the female wage 
\citep{Klasen2015,Bhargava2018}.
Second, women’s labor supply appears to be more negatively elastic to husbands' 
wages than positively elastic to own wages \citep{Bhargava2018}.

% [sex selection]

The introduction of sex selection allows parents to avoid giving birth to girls but 
increases the expected duration to the next birth.
Theory suggests that sex selection increases with lower desired fertility and with
higher parity for a given desired number of children \citep{Portner2015b}.
Sex selection is more widespread among better-educated than less-educated and among urban 
than rural women, which is consistent with lower desired fertility increasing the
use of sex selection 
\citep{das_gupta97,retherford03b,Guilmoto2009a,Portner2015b,Jayachandran2017}.
Better-educated and urban women also tend to live in households with higher income, which 
better enables them to access prenatal sex determination and lowers the relative costs of 
using sex selection and have long birth intervals.

The combination of rising incomes and continued son preference may lead to even longer 
spacing than expected from the income effect alone.
As women's education increases, their productivity in the production of offspring human 
capital also increases.
With relatively more boys born because of increased access to sex-selective 
abortions and the increasing income potential for (male) offspring, demand for 
better-educated women can increase, even if they do not participate in the labor market 
\citep{Behrman1999}.

If more and ``better'' parental attention per child results in higher child ``quality,'' 
we should expect longer birth intervals \citep{Zajonc1975,Zajonc1976,Razin1980}.
However, the evidence on spacing's effect on child quality measures such as IQ 
and education is mixed for developed countries and non-existing for developing countries
\citep{Powell1993,Pettersson-Lidbom2009,Buckles2012,Barclay2017}.
The exception is one crucial aspect of child quality---health and mortality---where
longer spacing does lead to better outcomes, although this relationship weakens with 
maternal education \citep{Whitworth2002,Conde-Agudelo2012,Molitoris2019}.

If parents realize that longer spacing leads to better child outcomes, we may even observe 
later conception, especially when, with sex selection, the next child is likely to be a boy.
Working in the opposite direction of this "delay" effect is that women with more education 
can, in principle, space their children closer together without substantially increasing 
child mortality risk. 
However, given the apparent lack of pressure to return to the labor force, this effect is 
likely small.

% Sanskritization

The increases in female educational attainment imply that access to education has 
expanded beyond the higher castes. 
One possible effect of the associated change in the composition of better-educated women 
is that this group's behavior would change.
However, ``Sanskritization'' implies that as lower castes females gain access to 
education and their husbands' income increases, they adopt higher caste 
norms such as stronger son preference and a retraction from the formal labor 
market \citep{Srinivas1956,Chen1995,Abraham2013,Chatterjee2018}.
The low and declining female labor force participation is an indication that this
process still operates.

% Summary / hypotheses

In summary, with substantial increases in husbands' income and a declining female labor 
force participation, I expect a push toward longer birth spacing over time, independent
of education levels, based on the income effects and the effects of spacing
on child outcomes.
Furthermore, I expect birth spacing to increase the most among the better-educated 
because their household income increases the most---even with declining female labor 
participation---and their use of sex selection.
Even with the substantial increase in the number of better-educated women, 
"Sanskritization" implies that the changing composition does not substantially change 
this group's use of sex selection.

@@

Direct information on the use of sex selection is not available, so I compare different 
periods, based on the changes in access and legality of prenatal sex 
determination in India.
Abortion has been legal in India since 1971.
Reports of sex determination appeared around 1982--83, and the number of clinics 
quickly increased \citep{Sudha1999,bhat06,Grover2006}.
In 1994, the Prenatal Diagnostic Techniques Act made determining and communicating 
the sex of a fetus illegal.%
\footnote{
Details about the act are at \href{http://pndt.gov.in/}{http://pndt.gov.in/}.
There is little evidence that the ban significantly affected sex ratios \citep{Das-Gupta2016}.
}
Although the use of sex selection increased even after 1994, we may have passed a 
turning point in its use \citep{Das_Gupta2009,Diamond-Smith2015}.
I, therefore, use four periods: 1972--1984, 1985--1994, 1995--2004, and 2005--2016.
The allocation of spells into periods is determined by when conception, and, therefore, 
decisions on sex selection, can begin. 
Hence, some spells cover two periods, which may bias downward the differences between the 
periods.

@@

The models are estimated separately for each of the four education group to reduce any
potential problem from including other variables as proportional since
women with different education levels have different hazard profiles
\citep{Whitworth2002,Bhalotra2008,Kim2010}.
I capture sex composition with dummy variables for the
possible combinations, ignoring the ordering of births.
Area of residence is a dummy variable for living in an urban area.
